Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-23 paper_enahpc14/Sec4:Evaluation.tex
creates subsubsubsection; adds new figures

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution}       and       \textit{energy-to-solution}.
Time-to-solution  refers   to  the  total  wall  clock   time  of  the
application execution time. Energy-to-solution is the amount of energy
spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
assessed  by sampling  the instantaneous  peak power  during execution
which  is then  averaged  and multiplied  by  the time-to-solution  to
determine energy-to-solution.   Whenever possible, multiple production
runs of COSMO-ART were  performed to illustrate the reproducibility of
the baseline, and quantify  the significant uncertainties in the power
measurement, as dictated by the available technology.

\subsection{Environment setup}
\label{subsec:4.2}
We have  chosen a state-of-the-art Intel's  third-generation Core (aka
Ivy  Bridge)  processing platform  (called  ``Monch'')  for our  power
measurements,  which is  slated to  stay in  service  without hardware
upgrade for  the duration of  the Exa2Green project.  In  principle at
least, this architecture  could be recreated or found  in an identical
configuration beyond the lifetime  of the project.  Since the baseline
benchmark can be reproduced within  an expected variance, and that the
baseline run configuration  can be used in all  future versions of the
code,  a fair comparison  will be  made between  the baseline  and the
milestone  versions  of COSMO-ART.   In  this  study, a  complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.   Finally a  full tracing  experiment is  conducted  on the
``Tintorrum''  cluster using  an integrated  framework, to  capture an
overall  power profile  at a  much finer  resolution and  to  get more
insights  into  the  MPI  blocking  and polling  influences  on  power
savings.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP  based cluster, utilised  by scientists who are  part of
the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS) of  ETH Zurich, is a dual-socket  eight-core Intel Sandy
Bridge EP based cluster used as Piz Daint pre-post processing cluster.
It is composed of 42  compute nodes and has 2 high-speed interconnects
based on FDR: the first is dedicated to the MPI traffic and the second
to  the storage  high speed  traffic.  The  2 login  nodes and  the 42
computes nodes consists in 11 twin-pair Intel E5-Series DALCO r2264i4t
2U  scalable compute modules.   Each module  contains 4  compute nodes
based on two  Intel Xeon E5-2670 processors operating  at 2.6 GHz base
clock speed,  themselves connected by a high  speed InfiniBand network
based  on  Mellanox  SX6036  managed  FDR switches,  with  a  56  Gb/s
speed. Each  core has 32 KB instruction  and 32 KB data  L1 caches and
256 KB  of L2 cache.  All  the 8 core share  a 20 MB L3  cache and the
platform has  64GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a full  rack of Pilatus constituted of  42 standard compute
nodes (pilatus[03-44]) was considered.

\subsubsection{Tintorrum (UJI)}
Tintorrum, which  is installed at the  Jaume I University  (UJI), is a
heterogeneous  cluster   composed  of  28  compute   nodes.   For  our
experiments only a subset of  these nodes were considered. This set is
composed  of 16 nodes,  each of  which includes  two Intel  Xeon E5645
hexa-core  processors  at  2.40   GHz  connected  via  Infiniband  QDR
(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
and the platform has 288 GB of DDR3 1333 MHz.

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimisation  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
computing speed regardless to  energy consumption. In our experiments,
Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
consumption of Xeon E5 by approximately $7\%$.

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART.eps}
  \caption{Time-to-solution and  energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures}
  \label{fig:3}
\end{figure}

>>>>

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
TTS refers to  the total wall clock time  of the application execution
time. ETS  is the amount of  energy spent to  achieve results.  Energy
consumption  is  assessed by  sampling  the  instantaneous peak  power
during execution which  is then averaged and multiplied  by the TTS to
determine  ETS.   Whenever   possible,  multiple  production  runs  of
COSMO-ART  were performed  to  illustrate the  reproducibility of  the
baseline,  and quantify  the  significant uncertainties  in the  power
measurement, as dictated by the available technology.

\subsection{Environment setup}
\label{subsec:4.2}

\subsubsection{HPC platforms}
We have  chosen a state-of-the-art Intel's  third-generation Core (aka
Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
power  measurements,  which  is  slated  to stay  in  service  without
hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Since
the baseline benchmark can  be reproduced within an expected variance,
and  that the baseline  run configuration  can be  used in  all future
versions  of the  code, a  fair comparison  will be  made  between the
baseline and  the milestone versions  of COSMO-ART.  In this  study, a
complementary  energy-to-solution benchmarking  comparison  is carried
out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
systems and known to be  more power consuming.  Finally a full tracing
experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
University (UJI) using an  integrated framework, to capture an overall
power profile at a much finer resolution and to get more insights into
the MPI blocking and polling influences on power savings.

\paragraph{CSCS - Monch} 
~\\
\vskip -1em
Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
based  cluster, utilised  by  scientists  who are  part  of the  Swiss
Platform     for      Advanced     Scientific     Computing     (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

\paragraph{CSCS - Pilatus} 
~\\
\vskip -1em
Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
cluster used as Piz Daint pre-post processing cluster.  It is composed
of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
the  first is  dedicated to  the  MPI traffic  and the  second to  the
storage high  speed traffic.   The 2 login  nodes and the  42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
speed, themselves  connected by a high speed  InfiniBand network based
on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
(pilatus[03-44]) was considered.

\paragraph{UJI - Tintorrum} 
~\\
\vskip -1em
Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
For our experiments only a subset of these nodes were considered. This
set is  composed of 16  nodes, each of  which includes two  Intel Xeon
E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
and the platform has 288 GB of DDR3 1333 MHz.

\subsubsection{Sofware environment}
The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
running on distributed multi-core systems only.  The software stack on
both CSCS  platforms was controlled using the  modules framework which
gives an easy and flexible mechanism  to access to all of the provided
compilers, tools  and applications.  For our  initial benchmarking, we
opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
delivering inferior performance.  In addition, we installed the MPICH2
implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
large  and  complex data  collections.   All  computes  nodes have  an
operating      system      based      on      GNU/Linux      featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
in this study as previous attempts  of its use revealed that it always
led to higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The reason for  that is  two-fold: (1)  it has  higher clock
frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
at  computing   speed  regardless   to  energy  consumption.   In  our
experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
reducing the energy consumption of Xeon E5 by approximately $7\%$.

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
  \caption{Time-to-solution and energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
  \caption{Time-to-solution and energy-to-solution on Tintorrum -
    Polling mode}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
  \caption{Time-to-solution and energy-to-solution on Tintorrum -
    Blocking mode}
  \label{fig:3}
\end{figure}

### 2014-04-23 paper_enahpc14/paper.tex
creates subsubsubsection; adds new figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[round, sort&compress, numbers]{natbib}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{units}
\usepackage{color}
\usepackage{xspace}
%
\DeclareRobustCommand\IPCClongname{}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evaluating the Performance and Energy Efficiency of COSMO-ART,
  a Coupled Numerical Weather Forecast and Chemical Transport Model.}
%\thanks{}
%\subtitle{}

@@

\begin{abstract}
  In  this paper  we present  the energy-to-solution  of  COSMO-ART on
  various platforms.   This model is  an extension of  the operational
  weather  forecast  model  of   the  German  weather  service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is analysed  by a profiling study to  determine hotspots and
  identify  critical  CPU paths.   Moreover,  we describe  measurement
  devices and energy-aware techniques  employed to evaluate the energy
  footprint of the considered application and to get detailed insights
  about power bottlenecks.  Our motivation is to improve corresponding
  code    sections   to    sustain   performance    while   minimizing
  energy-to-solution.   This  preliminary  work  sets  the  basis  for
  subsequent studies to tackle  challenges related to energy efficient
  high performance computing in the framework of the Exa2Green project
  (\url{http://exa2green.eu/}).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[round, sort&compress, numbers]{natbib}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{units}
\usepackage{color}
\usepackage{xspace}
%
\DeclareRobustCommand\IPCClongname{}
\setcounter{secnumdepth}{5}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evaluating the Performance and Energy Efficiency of the
  COSMO-ART model system}
%\thanks{}
%\subtitle{}

@@

\begin{abstract}
  In this  paper we investigate  the energy footprint  and performance
  profiling of COSMO-ART  on various HPC platforms.  This  model is an
  extension of  the operational weather  forecast model of  the German
  weather  service   (DWD),  developed  for  the   evaluation  of  the
  interactions of reactive gases  and aerosol particles with the state
  of atmosphere at the  regional scale.  Different measurement devices
  and energy-aware techniques are  described to evaluate both time and
  energy  to  solution  of  the  considered application  and  to  gain
  detailed  insights  into power  and  performance requirements.   Our
  motivation  is to  improve  corresponding code  sections to  sustain
  performance  while minimising energy-to-solution.   This preliminary
  work  sets the  basis for  subsequent studies  to  tackle challenges
  related  to  energy  efficient  high performance  computing  in  the
  framework of the Exa2Green project (\url{http://exa2green.eu/}).
